Active Learning for SAT Solver Benchmarking {\large (extended and revised version)}

### 2025-02-12 paper_jar/paper.tex
Add appendix

In our experiments, we work with the dataset of the SAT~Competition~2022 Anniversary Track~\cite{sat2022,zenodoanniversarytrack}.
This dataset consists of 5355 instances with corresponding runtimes of 28 sequential SAT solvers.
For predictions within our framework, we also use a database of 56 instance features\footnote{\url{https://benchmark-database.de/getdatabase/base_db}} from the Global Benchmark Database~(GBD)~\cite{GBD24paper}.
They include instance size features and node distribution statistics for several graph representations of SAT instances, and are primarily inspired by the SATzilla~2012 features described in~\cite{features}.
All features are numeric and have no missing values.
We drop 10 of the 56 features due to zero variance.
In total, the prediction models have access to 46 instance features and 27 out of 28 runtime features, excluding the respective new solver~$\hat{a}$.

In addition, we retrieve instance family information\footnote{\url{https://benchmark-database.de/getdatabase/meta_db}} to evaluate the composition of our sampled benchmarks.
Instance families consist of instances from the same application domain, such as planning, cryptography, etc., and are valuable for analyzing solver performance.
%
For hyperparameter tuning, we randomly sample \SI{10}{\%} of the complete set of 5355 instances with stratification regarding the instance family.
All instance families that are too \emph{small}, i.e., families with less than 10 instances, are put into one meta-family for stratification.
This \emph{tuning dataset} allows a more extensive exploration of the hyperparameter space.

@@

As a baseline, Table~\ref{tab:regression} shows the prediction performance of two regression models, i.e., a random forest and a multilayer perceptron.
We employ these models to estimate a new solver's raw runtime as the prediction target.
To this end, we use 46 instance features and 27 known solver runtimes as inputs for the predictions.
We repeat this evaluation for each solver as the prediction target and report the average and standard deviation of prediction performance over the 28 solvers.

@@

Next, we use the runtimes labels from this discretization procedure as prediction targets.
Table~\ref{tab:cluster-prediction} shows the prediction performance for the best classification models from Table~\ref{tab:timeout-prediction} and $k \in \lbrace 3, 4, 5 \rbrace$ clusters.
We train the prediction models either with 46 instance features and 27 known solver runtimes (second column) or additionally including 27 features representing the known $C_k$ cluster labels of other solvers (third column).

@@

For reproducibility, our source code and data are available on GitHub (see footnotes in Section~\ref{sec:intro}).
Our code is implemented in \textsc{Python} using \emph{scikit-learn}~\cite{scikit-learn} for making predictions, and \emph{gbd-tools}~\cite{GBD24paper} for retrieving SAT instances.

@@

>>>>

In our experiments, we work with the dataset of the SAT~Competition~2022 Anniversary Track~\cite{sat2022,zenodoanniversarytrack}.
This dataset consists of 5355 instances with corresponding runtimes of 28 sequential SAT solvers.
For predictions within our framework, we also use a database of 56 instance features from the Global Benchmark Database~(GBD)~\cite{IserJ24}.
\change{Refer to Appendix~\ref{sec:appendix-features} for a full list of all 56 features.}
They include instance size features and node distribution statistics for several graph representations of SAT instances, and are primarily inspired by the SATzilla~2012 features described in~\cite{features}.
All features are numeric and have no missing values.
We drop 10 of the 56 features due to zero variance.
In total, the prediction models have access to 46 instance features and 27 out of 28 runtime features, excluding the respective new solver~$\hat{a}$.

In addition, we retrieve instance family information\footnote{\url{https://benchmark-database.de/getdatabase/meta}} to evaluate the composition of our sampled benchmarks.
Instance families consist of instances from the same application domain, such as planning, cryptography, etc., and are valuable for analyzing solver performance.
For hyperparameter tuning, we randomly sample \SI{10}{\%} of the complete set of 5355 instances with stratification regarding the instance family.
All instance families that are too \emph{small}, i.e., families with less than 10 instances, are put into one meta-family for stratification.
This \emph{tuning dataset} allows a more extensive exploration of the hyperparameter space.

@@

As a baseline, Table~\ref{tab:regression} shows the prediction performance of two regression models, i.e., a random forest and a multilayer perceptron.
We employ these models to estimate a new solver's raw runtime as the prediction target.
To this end, we use 46 instance features \change{(Section~\ref{sec:exdesign:data})} and 27 known solver runtimes as inputs for the predictions.
We repeat this evaluation for each solver as the prediction target and report the average and standard deviation of prediction performance over the 28 solvers.

@@

Next, we use the runtimes labels from this discretization procedure as prediction targets.
Table~\ref{tab:cluster-prediction} shows the prediction performance for the best classification models from Table~\ref{tab:timeout-prediction} and $k \in \lbrace 3, 4, 5 \rbrace$ clusters.
We train the prediction models either with 46 instance features \change{(Section~\ref{sec:exdesign:data})} and 27 known solver runtimes (second column) or additionally including 27 features representing the known $C_k$ cluster labels of other solvers (third column).

@@

For reproducibility, our source code and data are available on GitHub (see footnotes in Section~\ref{sec:intro}).
Our code is implemented in \textsc{Python} using \emph{scikit-learn}~\cite{scikit-learn} for making predictions, and \emph{gbd-tools}~\cite{IserJ24} for retrieving SAT instances.

@@

%
% ---- Appendix ----
%

\newpage
\appendix
\section{Instance Features}
\label{sec:appendix-features}

\change{For predictions within our framework, we use a database of 56 instance features\footnote{\url{https://udopia.github.io/gbdc/doc/extractors/CNFBaseFeatures.html}} from the Global Benchmark Database~(GBD)~\cite{IserJ24}.
Table~\ref{tab:features} briefly summarizes all 56 features.
The placeholder \texttt{<n>} can take values within the range 1 to 9 and the placeholder \texttt{<dist>} is one of \texttt{mean}, \texttt{variance}, \texttt{min}, \texttt{max}, or \texttt{entropy}.
Accounting for all possible instantiations of \texttt{<n>} and \texttt{<dist>}, there are 56 features in total.}

\begin{table}[h]
    \centering
    \caption{
      \change{List of all 56 instance features, where \texttt{<dist>} is one of \texttt{mean}, \texttt{variance}, \texttt{min}, \texttt{max}, or \texttt{entropy}.}
    }
    \label{tab:features}
    \begin{tabular}{
      >{\arraybackslash}p{0.22\textwidth}
      >{\arraybackslash}p{0.21\textwidth}
      >{\arraybackslash}p{0.46\textwidth}
    }
      \toprule
      Category & Feature Name & Description \\
      \midrule
      Basic statistics & \verb|clauses| & Number of clauses \\
      & \verb|variables| & Number of variables \\
      \midrule
      Clause types & \verb|cls<n>| & Number of clauses with \verb|<n>| literals for \verb|<n>| in 1--9 \\
      & \verb|cls10p| & Number of clauses with 10 or more literals \\
      & \verb|horn| & Number of Horn clauses \\
      & \verb|invhorn| & Number of inverted Horn clauses \\
      & \verb|positive| & Number of positive clauses \\
      & \verb|negative| & Number of negative clauses \\
      \midrule
      Horn proximity & \verb|hornvars_<dist>| & Distribution of variable occurrence in horn clauses \\
      & \verb|invhornvars_<dist>| & Distribution of variable occurrence in inverted horn clauses \\
      \midrule
      Polarity balance & \verb|balancecls_<dist>| & Distributions of fractions of number of positive/negative literals in clauses \\
      & \verb|balancevars_<dist>| & Distributions of fractions of number of positive/negative literals of variables \\
      \midrule
      Variable-clause graph & \verb|vcg_vdegree_<dist>| & Variable degree distribution \\
      & \verb|vcg_cdegree_<dist>| & Clause degree distribution \\
      \midrule
      Variable graph & \verb|vg_degree_<dist>| & Degree distribution \\
      \midrule
      Clause graph & \verb|cg_degree_<dist>| & Degree distribution \\
      \bottomrule
    \end{tabular}
\end{table}
